---
title: "How to use poodleR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use poodleR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(poodleR)
library(data.table)
```

## Overview

`poodleR` estimates donor proportions in pooled sequencing experiments by combining:

- donor genotypes stored in a VCF
- pooled allele counts from `bam-readcount`

The typical workflow is:

1. Convert donor VCF genotypes to minor-allele dosages.
2. Convert pooled read counts to minor-allele frequency estimates.
3. Align both tables to the same SNPs in the same order.
4. Solve the constrained regression problem to estimate donor weights.

This vignette walks through every exported function using the example files that ship with the package.

## Example files

The package includes two compressed input files in `inst/extdata`:

```{r extdata}
bam_path <- system.file("extdata", "bam-readcount_example.tsv.gz", package = "poodleR")
vcf_path <- system.file("extdata", "vcf_example.vcf.gz", package = "poodleR")

bam_path
vcf_path
```

It also includes smaller saved objects in `data/` that are convenient for fast examples:

```{r data-objects}
data(A, package = "poodleR")
data(b, package = "poodleR")
data(b_estimate_dt, package = "poodleR")

dim(A)
dim(b)
dim(b_estimate_dt)
```

## `create_gt_dosages()`: convert the donor VCF

To demonstrate the genotype conversion step, load the bundled `vcf` object:

```{r create-gt-setup, eval = FALSE}
data(vcf, package = "poodleR")
```

```{r create-gt, eval = FALSE}
ma_dosages <- create_gt_dosages(vcf)
head(ma_dosages[, 1:5])
```

The result contains:

- one row per retained SNP
- one numeric donor column per sample
- an `rn` column storing `CHROM_POS_REF`

Rows with duplicated identifiers or missing donor genotypes are removed during this step.

## `estimate_b_from_bam_readcount()`: convert pooled read counts

The lower-level read-count conversion function expects:

- the first ten bam-readcount columns
- an `ALT` column from a matching VCF subset
- a `CHROM_POS_REF` key shared by both objects

The saved example objects let us demonstrate the format without repeating the full import pipeline:

```{r estimate-b}
head(b_estimate_dt)
summary(b_estimate_dt$b_estimate)
```

If you want to build this table yourself from raw objects, the pattern is:

```{r estimate-b-from-objects, eval = FALSE}
data(bam_readcount, package = "poodleR")
data(vcf, package = "poodleR")

colnames(bam_readcount) <- c("CHROM", "POS", "REF", "TOTAL_READS", "DEL", "A", "C", "G", "T", "N")
bam_readcount$CHROM_POS_REF <- paste(bam_readcount$CHROM, bam_readcount$POS, bam_readcount$REF, sep = "_")

vcf_reduced <- as.data.table(vcf@fix[, 1:5])
vcf_reduced$CHROM_POS_REF <- paste(vcf_reduced$CHROM, vcf_reduced$POS, vcf_reduced$REF, sep = "_")
vcf_reduced <- vcf_reduced[ALT %in% c("A", "C", "G", "T")]
vcf_reduced <- vcf_reduced[vcf_reduced$CHROM_POS_REF %in% bam_readcount$CHROM_POS_REF, ]
bam_readcount <- bam_readcount[bam_readcount$CHROM_POS_REF %in% vcf_reduced$CHROM_POS_REF, ]

b_estimate_table <- estimate_b_from_bam_readcount(bam_readcount, vcf_reduced)
head(b_estimate_table)
```

## `calculate_nonRef_vector_and_genotype_df()`: do the full preprocessing

For most workflows, the easiest entry point is the combined helper:

```{r calculate-nonref, eval = FALSE}
prepared <- calculate_nonRef_vector_and_genotype_df(
  bam_readcount_path = bam_path,
  vcf_path = vcf_path,
  min_read_overlap = 5
)
```

It returns two aligned tables:

```{r prepared-objects, eval = FALSE}
names(prepared)
nrow(prepared$b_estimate_table)
nrow(prepared$ma_dosages)
head(prepared$b_estimate_table)
head(prepared$ma_dosages[, 1:5])
```

You can also optionally save the filtered outputs to disk with:

- `b_estimate_path`
- `genotype_minor_allele_dos_path`
- `variants_retained_path`

## `is_identical_genotype()`: inspect uninformative SNP rows

Before solving the regression, `estimate_weights()` removes SNP rows where every donor has the same genotype. You can inspect that filter directly:

```{r identical-genotypes}
donor_cols <- setdiff(colnames(A), "rn")
identical_rows <- is_identical_genotype(as.matrix(A[, donor_cols, with = FALSE]))
table(identical_rows)
```

Rows flagged as `TRUE` do not help distinguish donors and are excluded from the regression.

## `mod_lsqlincon()`: constrained least squares

The regression solver can also be called directly:

```{r mod-lsqlincon, eval = FALSE}
C <- matrix(c(0, 1,
              0.5, 0,
              1, 0.5), ncol = 2, byrow = TRUE)
d <- c(0.2, 0.5, 0.8)

mod_lsqlincon(
  C = C,
  d = d,
  Aeq = matrix(1, nrow = 1, ncol = 2),
  beq = 1,
  lb = 0,
  ub = 1
)
```

This is the lower-level solver used by `estimate_weights()`.

## `estimate_weights()`: estimate donor proportions

Once you have aligned genotype dosages and pooled allele frequencies, donor deconvolution is one function call:

```{r estimate-weights}
weights <- estimate_weights(
  b = b_estimate_dt$b_estimate,
  A = A
)

weights
sum(weights)
```

The result is a named numeric vector whose values are constrained to lie between 0 and 1 and sum to 1.

## End-to-end usage

In practice, most analyses will look like this:

```{r full-workflow, eval = FALSE}
prepared <- calculate_nonRef_vector_and_genotype_df(
  bam_readcount_path = bam_path,
  vcf_path = vcf_path,
  min_read_overlap = 5
)

weights <- estimate_weights(
  b = prepared$b_estimate_table$b_estimate,
  A = prepared$ma_dosages
)

sort(weights, decreasing = TRUE)
```

That gives you a complete pooled-donor deconvolution workflow from the shipped example files.
